---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Alex Lopez"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

#load packages
library(tidyverse)
library(agricolae)   
library(here)

#check wd
here()

#import data
raw.ntl.lter <- 
  read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
           stringsAsFactors = TRUE)

#format dates
raw.ntl.lter$sampledate <- as.Date(raw.ntl.lter$sampledate, format = "%m/%d/%y")

#2

#set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        axis.title.x = element_text(size = 10),               
        axis.title.y = element_text(size = 10),
        legend.position = "right",
        legend.title = element_text(size = 10),   
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 12))
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant linear relationship between mean July temperature and depth.
Ha: There is a significant linear relationship between mean July temperature and depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

ntl.lter.july <- raw.ntl.lter %>% 
  filter(month(sampledate) == 07) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5

tempvsdepth <- ggplot(ntl.lter.july, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', col = 'purple') +
  labs(x = "Depth (m)", y = "Temperature (°C)", 
       title = "Temperature vs. Depth in July") +
  ylim(0, 35)

print(tempvsdepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer:
The figure suggests that there is an inverse relationship between depth and temperature. The distribution of points suggests that this trend is not strictly linear, especially when depth is greater than 5m, leading us to believe that a linear model might not be the best fit for this data.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

lin.reg <- lm(data = ntl.lter.july, temperature_C ~ depth)
summary(lin.reg)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
According to the regression analysis, depth can be a predictor of temperature. Depth explains approximately 73.9% (R-squared value of 0.7387) of variability in temperature. This finding is based on 9726 degrees of freedom. The statistical significance is less than 0.05, which means we can reject the null hypothesis in favor of the alternative, indicating a signifcant relationship between depth and temperature. For every 1m change in depth, temperature is predicted to decrease by -1.94621 degrees Celsius.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

mlt.lin.reg <- lm(data = ntl.lter.july, temperature_C ~ year4 + daynum + depth)

step(mlt.lin.reg)

#10

summary(mlt.lin.reg)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
According to the AIC, the recommended set of explanatory variables best suited to predict temperature includes year4, daynum, and depth, because if we remove any of these variables, AIC increases. The smaller the AIC, the better. According to the regression analysis, this model explains approximately 74.1% (R-squared value of 0.7411) of the observed variance. This is a slight improvement over the model using only depth as the explanatory variable, an increase to 0.7411 from 0.7387.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#ANOVA as aov
july.temps.anova <- aov(data = ntl.lter.july, temperature_C ~ lakename)
summary(july.temps.anova)

#ANOVA as lm
july.temps.anova2 <- lm(data = ntl.lter.july, temperature_C ~ lakename)
summary(july.temps.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The null hypothesis in this case states that all lakes have the same mean temperature. The alternative hypothesis states that the mean temperature is not the same for all lakes. According to the ANOVA test, there is a highly signifcant p-value (< 0.05) associated with 'lakename' in the ANOVA model and a highly signifcant p-value (< 0.05) in the F-test of the linear model. Therefore, we can reject the null hypothesis in favor of the alternative, leading us to conclude that there is indeed a significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(ntl.lter.july, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_smooth(method = "lm", se = F) +
  labs(x = "Depth (m)", y = "Temperature (°C)", 
       title = "Temperature vs. Depth in July (by Lake)", color = "Lake Name") +
  ylim(0, 35)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

#post-hoc test
TukeyHSD(july.temps.anova)

#extract groupings for pairwise relationships
july.temps.groups <- HSD.test(july.temps.anova, "lakename", group = TRUE)
july.temps.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Statistically speaking, only Paul Lake has the same mean temperature as Peter Lake. After extracting the groupings for pairwise relationships in a plot, we can identify which lakes have the same mean temperature, statistically speaking, by looking at those that share the same letters. In our case, only Paul Lake is in the same letter grouping as Peter Lake.
After running the Tukey HSD function and looking at the p-values for all the pariwise relationships, we can observe that there is not one lake with a mean temperature that is statistically distinct from all the other lakes. Each lake has at least one other lake with which, together, their difference in means is not statistically distinct, because the p-value is greater than 0.05 for that pairing.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
We could conduct a two-sample t-test to compare the mean temperatures of Peter Lake and Paul Lake. The null hypothesis would state that the mean temperatures of Peter Lake and Paul Lake are equal, and the alternative hypothesis would state that they have different mean temperatures.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

#wrangle july data
crampton.and.ward <- ntl.lter.july %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

#run test
temps.twosample <- 
  t.test(crampton.and.ward$temperature_C ~ crampton.and.ward$lakename)
temps.twosample

```

>Answer: 
The null hypothesis states that the difference between the mean temperatures for Crampton Lake and Ward Lake is zero, and the alternative hypothesis states that the difference between the mean temperatures of both lakes is not zero. According to the t-test, the p-value is 0.2649, greater than 0.05, so we fail to reject the null hypothesis. In other words, the mean temperatures for Crampton Lake and Ward Lake are statistically the same.
